disaggregation: An \proglang{R} Package for Bayesian Spatial Disaggregation Modelling

### 2019-12-10 article.Rnw
Dan's comments

For the disaggregation model we model the rate at pixel level, with the likelihood for then observed data then given by aggregating these pixel level rates. The rate in pixel $j$ of polygon $i$ at location $s_{ij}$ is given by:
%
\begin{equation} \label{eq:lp}
\textrm{link}(\textrm{rate}_{ij}) = \beta_0 + \beta X_{ij} + \textrm{GP}(s_{ij}) + u_i 
\end{equation}
%
where $\beta$ are regression coefficients, $X_{ij}$ are covariate values, GP is a Gaussian random field and $u_i$ is a polygon-specific iid effect. The user defined link function is typically log, identity and logit for Poisson, Normal and Binomial likelihoods, respectively  The Gaussian random field has a Matern covariance function parameterised by $\rho$, the range (approximately the distance beyond which correlation is less than 0.1), and $\sigma$, the marginal standard deviation. As we are working in a Bayesian setting, each of the model parameters and hyperparameters are given a prior, which is discussed later.

@@

In order to know which pixels (i.e. which rows) are contained in each polygon, a matrix is constructed that contains the start and end pixel index for each polygon (\code{getStartendindex} function). Additionally, an INLA mesh is built to use for the spatial field (\code{build_mesh} function). The \code{mesh.args} argument allows the user to supply a list of INLA mesh parameters to control the mesh used for the spatial field. The mesh can take a while to construct, so to prepare the data without buidling the mesh the user can set the \code{makeMesh} flag to FALSE. However, it would then not be possible to fit the disaggregation model without the mesh.

If there are any NAs in the response or covariate data within the polygons the \code{prepare_data} method will return an error. This can be dealt with using the \code{na.action} flag, which is automatically off.  Ideally the NAs in the data would be dealt with by the user beforehand, however, setting na.action = TRUE will automatically deal with NAs. It removes any polygons that have NAs as a response, sets any aggregation pixels with NA to zero and sets covariate NAs pixels to the median value for the that covariate.

@@

%
<<subplotdata, fig.cap='Maps of Madagascar showing the data used in the disaggregation model. These plots are produced when calling the plot function on the disag.data object.', fig.subcap=c('Number of malaria cases in each administrative unit', 'Maps of the four covariates used in the model', 'inla.mesh object that will be used to make the spatial field'), out.width='0.5\\linewidth', echo=FALSE, fig.ncol = 2, dev = 'png', fig.width = 4, fig.height = 5, fig.pos = 'H'>>=
data_plots$polygon
data_plots$covariates
INLA::plot.inla.mesh(dis_data$mesh, main = 'INLA mesh for spatial field')
@
%

@@

\begin{figure}[!h]
\centering
%
<<plotmodel, fig.width = 10, fig.height = 6>>=
plot(fitted_model)
@
%
\caption{\label{fig:fitmodel} Plot summarising the results of the fitted model. These plots are produced when calling the \code{plot} function on the \code{fit.result} object. The fixed effects plot (left) shows the fitted parameter values with uncertainty estimation for all the fixed effects in the model. The in-sample performance plot (right) shows the predicted incidence rate values for each polygon in the data against the observed values for that polygon in the data.}
\end{figure}

The \code{fit_model} function returns an object of class \code{fit.result}, which is designed to be used directly in the \code{predict} function.
Therefore, now we have fitted the model, we are ready to predict the malaria incidence rate across Madagascar.

@@

%
<<subpredictmean, fig.cap='Maps of Madagascar showing the fine-scale predictions from the disaggregation model. These maps are produced when calling the plot function on the predictions object. The maps show the (a) mean prediction of malaria incidence rate, (b) covariate contribution to the linear predictor, and (c) field contribution to the linear predictor', fig.subcap=c('', '', ''), out.width='0.33\\linewidth', echo=FALSE, dev = 'png', fig.width = 4, fig.height = 6, fig.pos = 'H'>>=
mean_predictions$mean
mean_predictions$covariates
mean_predictions$field
@
%

@@

Disaggregation modelling, which involves fitting models at fine-scale resolution using areal data over heterogenous regions, has become widely used in fields such as epidemiology and ecology. The \pkg{disaggregation} package implements Bayesian spatial disaggregation modelling with a simple, easy to use \proglang{R} interface. The package includes simple data preparation, fitting and prediction functions that allow some user defined model flexibility. In this paper we have presented an application of the package, predicting malaria incidence rate across Madagascar from aggregated count data and environmental covariates.

>>>>

For the disaggregation model we model the rate at pixel level, with the likelihood for the observed data given by aggregating these pixel level rates. The rate in pixel $j$ of polygon $i$ at location $s_{ij}$ is given by:
%
\begin{equation} \label{eq:lp}
\textrm{link}(\textrm{rate}_{ij}) = \beta_0 + \beta X_{ij} + \textrm{GP}(s_{ij}) + u_i 
\end{equation}
%
where $\beta$ are regression coefficients, $X_{ij}$ are covariate values, GP is a Gaussian random field and $u_i$ is a polygon-specific iid effect. The user-defined link function is typically log, identity and logit for Poisson, Normal and Binomial likelihoods, respectively.  The Gaussian random field has a Matern covariance function parameterised by $\rho$, the range (approximately the distance beyond which correlation is less than 0.1), and $\sigma$, the marginal standard deviation. As we are working in a Bayesian setting, each of the model parameters and hyperparameters are given a prior, which is discussed later.

@@

In order to know which pixels (i.e. which rows) are contained in each polygon, a matrix is constructed that contains the start and end pixel index for each polygon (\code{getStartendindex} function). Additionally, an INLA mesh is built to use for the spatial field (\code{build_mesh} function). The \code{mesh.args} argument allows the user to supply a list of INLA mesh parameters to control the mesh used for the spatial field. The mesh can take several minutes to construct, so to prepare the data without buidling the mesh the user can set the \code{makeMesh} flag to FALSE. However, it would then not be possible to fit the disaggregation model without the mesh.

If there are any NAs in the response or covariate data within the polygons the \code{prepare_data} method will return an error. This can be dealt with using the \code{na.action} flag, which is automatically off.  Ideally the NAs in the data would be dealt with by the user beforehand, however, setting na.action = TRUE will automatically deal with NAs. It removes any polygons that have NAs as a response, sets any aggregation pixels with NA to zero and sets covariate NAs pixels to the median value for that covariate across all polygons.

@@

%
<<subplotdata, fig.cap='Maps of Madagascar showing the data used in the disaggregation model. These plots are produced when calling the plot function on the disag.data object. The plots show: (a) number of malaria cases in each administrative unit, (b) maps of the four covariates used in the model; and (c) inla.mesh object that will be used to make the spatial field.', fig.subcap=c('', '', ''), out.width='0.5\\linewidth', echo=FALSE, fig.ncol = 2, dev = 'png', fig.width = 4, fig.height = 5, fig.pos = 'H'>>=
data_plots$polygon
data_plots$covariates
INLA::plot.inla.mesh(dis_data$mesh, main = 'INLA mesh for spatial field')
@
%

@@

\begin{figure}[!h]
\centering
%
<<plotmodel, fig.width = 13, fig.height = 6>>=
plot(fitted_model)
@
%
\caption{\label{fig:fitmodel} Plot summarising the results of the fitted model. These plots are produced when calling the \code{plot} function on the \code{fit.result} object. The fixed effects plot (left) shows the fitted parameter values with uncertainty estimation for all the fixed effects in the model. The in-sample performance plot (right) shows the predicted incidence rate values for each polygon in the data against the observed values for that polygon in the data.}
\end{figure}

The \code{fit_model} function returns an object of class \code{fit.result}, which is designed to be used directly in the \code{predict} function.
Therefore, now that we have fitted the model, we are ready to predict the malaria incidence rate across Madagascar.

@@

%
<<subpredictmean, fig.cap='Maps of Madagascar showing the fine-scale predictions from the disaggregation model. These maps are produced when calling the plot function on the predictions object. The maps show the (a) mean prediction of malaria incidence rate, (b) covariate contribution to the linear predictor, and (c) field contribution to the linear predictor.', fig.subcap=c('', '', ''), out.width='0.33\\linewidth', echo=FALSE, dev = 'png', fig.width = 4, fig.height = 6, fig.pos = 'H'>>=
mean_predictions$mean
mean_predictions$covariates
mean_predictions$field
@
%

@@

Disaggregation modelling, which involves fitting models at fine-scale resolution using areal data over heterogenous regions, has become widely used in fields such as epidemiology and ecology. The \pkg{disaggregation} package implements Bayesian spatial disaggregation modelling with a simple, easy to use \proglang{R} interface. The package includes simple data preparation, fitting and prediction functions that allow some user-defined model flexibility. In this paper we have presented an application of the package, predicting malaria incidence rate across Madagascar from aggregated count data and environmental covariates.
